---
title: Calculate average from schemes
df-print: kable
---

## Introduction

This document generates a set of "average mitigation assumptions" for each of
96 Types of Potentially Mitigatable Activity (TPMA).

The results are shown in the final table of the document.
They are also made available as an xlsx file.

The document also shows some intermediate data objects for reference, and
documents the assumptions made in the code in order to arrive at the final
table, for checking and transparency.

## Assumptions and decisions made about the process

The assumptions made are, in summary:

1. Only model runs tagged with any of the run stages listed below are used
2. A maximum of 1 model run per scheme is included, based on a preference order
  of run_stage tags
3. Only `activity_avoidance` and `efficiencies` intervals are included
4. All TPMAs are included, by default. The exception is a small number of TPMAs
  labelled `bads_*`, which are excluded due to not matching the list provided in
  a reference `mitigators.json` file.
5. Intervals for all types of activity (inpatient, outpatient, aae) are
  included, by default.
6. `p10`, `p50` and `p90` values are reported, taken from the distributions
  created from all interval values available for each TPMA, across all selected
  model runs. Currently some intervals are only available in a small number of
  model runs. It would be possible to filter out TPMAs from the final table
  where the summary values are taken from a distribution based on small numbers
  of model runs (eg 4 or fewer?).
7. Out-of-range interval values provided by schemes are excluded. This includes
  4 cases where schemes have provided p10/p90 values greater than 1, and one
  case where a p10 value is given as 0.1 (ie 90% reduction in activity), which
  is thought to be a user error.
8. `p10`, `p50` and `p90` values are printed in the final table rounded to 3dp,
  for presentational reasons. It is simple to provide these instead without
  rounding if desired.

Where these assumptions are not in line with what is required, it is simple to
make changes to the process.
Please contact Fran Barton with details of any amendments, or any questions.

## Run stage tags in priority order

Only runs with the following tags can be included, and for each scheme only 1
run is chosen, based on the order of the following list (top priority at the
top):

* `final_report_ndg2`
* `final_report_ndg3`
* `final_report_ndg3_rd`
* `final_report_ndg1`
* `intermediate_ndg2`
* `intermediate_ndg3`
* `intermediate_ndg1`
* `initial_ndg2`
* `initial_ndg3`
* `initial_ndg1`

If a scheme has two tagged runs, one tagged `final_report_ndg2` and one
`intermediate_ndg2`, only the former will be chosen.

## Data process

### Initial data

We retrieve the params data for all tagged runs, from a `pins` board, and a
csv file with a list of "mitigators" and their names.

```{r get pinned params data}
#| cache: true
#| message: false
board <- pins::board_connect(auth = "envvar")

nhp_tagged_runs_meta <- board |>
  pins::pin_read("matt.dray/nhp_tagged_runs_meta")

params_data_init <- board |>
  pins::pin_read("matt.dray/nhp_tagged_runs_params")

tpma_lookup <- azkit::get_container("supporting-data") |>
  azkit::read_azure_csv("mitigator-lookup.csv", col_types = "-ccc-----") |>
  dplyr::rename(
    change_factor = "mitigator_type",
    strategy = "mitigator_variable",
    strategy_name = "mitigator_name"
  )
```

We begin with data for `r length(params_data_init)` runs.

`tpma_lookup` provides a list of `r nrow(tpma_lookup)` "strategies".
Later we will use this to ensure that only intervals for strategies included
in this list are used in the final data output.

```{r filter to preferred params per dataset}
run_stages <- c(
  "final_report_ndg2",
  "final_report_ndg3",
  "final_report_ndg3_rd",
  "final_report_ndg1",
  "intermediate_ndg2",
  "intermediate_ndg3",
  "intermediate_ndg1",
  "initial_ndg2",
  "initial_ndg3",
  "initial_ndg1"
)

preferred_runs <- nhp_tagged_runs_meta |>
  dplyr::filter(dplyr::if_any("run_stage", \(x) x %in% {{ run_stages }})) |>
  dplyr::slice_min(
    match(.data[["run_stage"]], {{ run_stages }}),
    by = "dataset"
  ) |>
  dplyr::select(!"file")

preferred_runs
```

The initial set of `r nrow(nhp_tagged_runs_meta)` is now filtered down to
`r nrow(preferred_runs)` preferred runs.

### Summary of selected runs to be included

```{r filter params data to preferred runs per dataset only}
preferred_run_names <- preferred_runs |>
  glue::glue_data("{dataset}_{scenario}_{create_datetime}")

length(preferred_run_names) # 24

params_data <- params_data_init |>
  purrr::keep_at(preferred_run_names)

# check all names are found
length(params_data) == length(preferred_run_names)

core_elements <- c(
  "dataset",
  "scenario",
  "start_year",
  "end_year",
  "app_version",
  "create_datetime"
)
params_data_tbl_summary <- params_data |>
  purrr::map(\(x) purrr::keep_at(x, core_elements)) |>
  purrr::map(tibble::as_tibble_row) |>
  purrr::list_rbind() |>
  dplyr::mutate(dplyr::across(tidyselect::ends_with("year"), as.integer))

```

This table is not used in the further calculation below, but is shown here for
information.

```{r display summary table}
params_data_tbl_summary |>
  dplyr::arrange(.data[["create_datetime"]])
```

### Extract intervals from selected params

Here we take the `activity_avoidance` and `efficiencies` elements of the params
data only.
For each of the 24 runs, and for each "strategy", we extract the provided `p10`
and `p90` values and pivot these into columns in a data frame.

```{r extract intervals}
create_intervals_tbl <- function(lst) {
  lst |>
    purrr::map_depth(2, \(x) list(x[["interval"]])) |>
    purrr::imap(\(x, y) dplyr::bind_cols(type = y, tibble::as_tibble_row(x))) |>
    purrr::list_rbind() |>
    tidyr::pivot_longer(
      !"type",
      names_to = "strategy",
      values_drop_na = TRUE
    ) |>
    tidyr::unnest_wider("value", names_sep = "_") |>
    dplyr::rename(p10 = "value_1", p90 = "value_2")
}


combined_intervals_tbl <- params_data |>
  purrr::map(\(x) purrr::keep_at(x, c("activity_avoidance", "efficiencies"))) |>
  purrr::map_depth(2, create_intervals_tbl) |>
  purrr::map(\(x) purrr::list_rbind(x, names_to = "change_factor")) |>
  purrr::list_rbind(names_to = "run_label")

```

The `combined_intervals_tbl` data frame contains
`r nrow(combined_intervals_tbl)` rows.

We then filter `combined_intervals_tbl` against the `tpma_lookup` table
that was read in at the top of the document.
Intervals are only retained if both their `change_factor` (activity_avoidance or
efficiencies) and `strategy` fields are matched.

```{r counts of combinations - number of runs as basis for mixtures}

tidy_intervals_tbl <- combined_intervals_tbl |>
  # only include "approved" list of strategies (excludes "bads_*"))
  dplyr::inner_join(tpma_lookup, c("change_factor", "strategy")) |>
  # Rationale for these exclusions is here:
  # https://github.com/The-Strategy-Unit/nhp_analysis/blob/cbf7e809419b516d6d8e2977fbf5d5c950263b7d/2025-02-28-wp1a-communties/wp1a.qmd#L159
  dplyr::filter(
    dplyr::if_any("strategy", \(x) !grepl("^virtual", x)) |
      dplyr::if_any(c("p10", "p90"), \(x) !is.na(x))
  ) |>
  dplyr::filter(
    !grepl("^RH5", .data[["run_label"]]) |
      .data[["strategy"]] != "virtual_wards_activity_avoidance_ari"
  ) |>
  dplyr::filter(dplyr::if_all(c("p10", "p90"), \(x) {
    dplyr::between(x, 0, 1) # 4 rows are set at above 1
  }))

counts_summary <- tidy_intervals_tbl |>
  dplyr::summarise(
    n_runs_included = dplyr::n(),
    .by = c("type", "change_factor", "strategy", "strategy_name")
  ) |>
  dplyr::arrange(dplyr::desc(.data[["n_runs_included"]]))

# counts_summary

```

The `counts_summary` table above shows that in all cases the set of intervals
for a given "strategy" is less than the potential maximum of 24 (which would be
where intervals were provided in every selected run), and in some cases the set
is very small; for example, the intervals being provided in only 1 of the 24
runs.

### Create distributions and compile final table

After filtering out some rogue values, we now have a `tidy_intervals_tbl` table
with `r nrow(tidy_intervals_tbl)` rows.


### Standardise intervals

Because the provided intervals (`p10` and `p90` values) we are using were
supplied as projections of activity avoidance / efficiencies over different time
spans (eg some were 2019-2036, others 2023-2041, for example), we need to
standardise each set of intervals to an annualised rate, before creating
the mixture distributions, and then multiply up our results to a standard time
period.


```{r standardise projections}
run_lookup <- params_data_tbl_summary |>
  dplyr::select(!"app_version") |>
  tidyr::unite(run_label, c("dataset", "scenario", "create_datetime"))

annualise <- \(x, p) `^`(x, (1 / p))

annualised_intervals_tbl <- tidy_intervals_tbl |>
  dplyr::left_join(run_lookup, "run_label") |>
  dplyr::mutate(
    period = .data[["end_year"]] - .data[["start_year"]],
    .keep = "unused"
  ) |>
  dplyr::mutate(
    annualised_p10 = annualise(.data[["p10"]], .data[["period"]]),
    annualised_p90 = annualise(.data[["p90"]], .data[["period"]]),
    .keep = "unused"
  )

```


Now we use a function to generate a distribution across all `p10` and `p90`
values for each strategy.
We then apply a "reinflation" function that converts the annualised p10 and p90
values back to quantities that would apply over, in this case, 5 years or 7
years.
(This raises the annualised values to the power of 5 and 7
respectively).
We then report back the `p10`, `p50` and `p90` values for each of those
reinflated distributions.


```{r create mixtures function}
create_mixtures <- function(intervals_tbl, colnames_vec) {
  intervals_tbl |>
    dplyr::mutate(
      mu = (.data[["annualised_p10"]] + .data[["annualised_p90"]]) / 2,
      sigma = (.data[["annualised_p90"]] - .data[["mu"]]) /
        qnorm(p = 0.90, mean = 0, sd = 1),
      dplyr::across("sigma", \(x) dplyr::if_else(x == 0, 0.0001, x)),
      dist = purrr::map2(.data[["mu"]], .data[["sigma"]], \(m, s) {
        distr::Truncate(distr::Norm(mean = m, sd = s), lower = 0, upper = 1)
      })
    ) |>
    dplyr::summarise(dplyr::across("dist", list), .by = colnames_vec) |>
    dplyr::mutate(
      mixture = purrr::map(.data[["dist"]], \(x) {
        distr::UnivarMixingDistribution(Dlist = x)
      }),
      .keep = "unused"
    )
}

```


```{r reinflation function}
reinflate_intervals <- function(tbl, x, exp) {
  dplyr::mutate(tbl, "{x}_{exp}yr" := `^`(.data[[x]], exp))
}
reinflate_5yr <- purrr::partial(reinflate_intervals, ... = , exp = 5)
reinflate_7yr <- purrr::partial(reinflate_intervals, ... = , exp = 7)
```

```{r create mixtures across all runs then reinflate values}
#| warning: false

pcols <- c("p10", "p50", "p90")

final_table <- annualised_intervals_tbl |>
  create_mixtures(c("type", "change_factor", "strategy", "strategy_name")) |>
  dplyr::mutate(
    p10 = purrr::map_dbl(.data[["mixture"]], \(m) m@q(p = 0.1)),
    p50 = purrr::map_dbl(.data[["mixture"]], \(m) m@q(p = 0.5)),
    p90 = purrr::map_dbl(.data[["mixture"]], \(m) m@q(p = 0.9)),
    .keep = "unused"
  ) |>
  purrr::reduce(pcols, reinflate_5yr, .init = _) |>
  purrr::reduce(pcols, reinflate_7yr, .init = _) |>
  dplyr::select(!tidyselect::all_of(pcols)) |>
  dplyr::mutate(dplyr::across("type", \(x) {
    forcats::fct(x, c("ip", "op", "aae"))
  })) |>
  dplyr::arrange(dplyr::pick(c("type", "change_factor")))


```


NB if it is preferable to produce the table below pivoted to a long format,
with three value columns p10, p50, p90, and a period identifier column (5yr
projection or 7yr projection), this can easily be provided instead.

```{r display final table}
final_table |>
  dplyr::mutate(dplyr::across(tidyselect::matches("^p[159]0"), \(x) {
    round(x, 3)
  }))
```


```{r export data}

final_table |>
  readr::write_rds("avg_intervals_from_selected_tagged_runs.rds")

final_table |>
  openxlsx2::write_xlsx("avg_intervals_from_selected_tagged_runs.xlsx")

```
